Probabilities and propensities: conversion experiences in online grocery shopping

### 2017-07-17 reconstruction.rmd
starting reconstruction

---
title: "Probabilities and propensities: conversion experiences in online grocery shopping"
author: Adrian Mackenzie^[University of Somewhere] 
date: "`r format(Sys.time(), '%B %Y')`"
fontsize: 12pt
citation_package: biber
latex_engine: xelatex
bibliography: /home/mackenza/ref_bibs/uni.bib
csl: "sage-harvard.csl"
link-citations: yes
header-includes:
    - \usepackage{setspace}
    - \doublespacing
abstract: 
    "Recommender systems attempt to transform or convert the temporal and agential fabric of social ordering by substituting probabilistic calculation for social practices. This paper investigates practices of ordering in the setting of shopping lists and  online grocery orders. It reconstructs an account of how a recommender system suggests a small number of grocery items of personal relevance to each of the millions of online grocery shoppers in  a major UK supermarket chain. Although all shopping lists mix past and future actions, actual and potential events, and normative and actual values, their re-writing by recommender systems adds a new operational dynamic.The paper explores how in the context of recommender systems the constitutive incompleteness of shopping lists, their propensity to expand or change, becomes more important than their capacity to inscribe social order. This suggests that the conversion of lists from finite ordering to indeterminate propensity in recommender systems requires new ways of conceptualising social order."
keywords: prediction, probability, order, lists, recommender system, shopping
---

```{r setup2, cache=FALSE, include=FALSE}
library(knitr)
library(tint)
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
output <- opts_knit$get("rmarkdown.pandoc.to")
opts_chunk$set(warning=FALSE)
opts_chunk$set(message=FALSE)
opts_chunk$set(echo=FALSE)
opts_chunk$set(error=FALSE)
opts_chunk$set(cache=TRUE)
```

# Probabilities and propensities: conversion experiences in online grocery shopping

Online grocery shopping at Tesco, a large supermarket chain based in the UK, includes recommendations for further grocery purchases under the title of 'Have you forgotten?'. When Tesco customers shop for groceries online, a list of five recommendations appear at the checkout stage. The recommendations are the product of a recommender system, an important category of operational device in big data (see for instance [@Striphas_2015] for analysis of Netflix recommendation; [@Morris_2015]  or [@Seaver_2015] for an account of music recommendations]).  The question 'have you forgotten?' is followed by a list of some  grocery items that could have been or are usually on a shopping list. The title of the suggestions is a bit misleading: the recommender system, as we will see, is not concerned with forgetting, with the many slips and oversights associated to which shopping is prone,  but rather with items that customers had not selected perhaps because they had never thought of buying them in the first place.

The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco. For shoppers, as Daniel Miller has argued [@Miller_2012],  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between ideals of health and commitments to thrift). It can also be seen  as a sacrificial ritual that re-instates social order in various ways.

The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of these logistical flows, infrastructural orderings, cultural values and  lively negotiations.  It is an intersectional ordering device that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

```{r random-basket, echo=FALSE, cache=TRUE, fig.cap='Random basket'}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)
csvs = list.files('data', pattern = '*.csv',full.names=TRUE)
df = bind_rows(sapply(csvs, read_tsv))
kable(sample_n(df, size=5)$name, col.names='item', caption='A random list of Tesco grocery items')
```

```{r shopping-hand, fig.cap='A hand-written shopping list'}
include_graphics('figure/1_shopping.png')
```

The items on Tesco's 'Have You Forgotten?' list are recommendations. They are only a few of them: five. The question is: which five items and in what order? In terms of content, there are many similarities between shopping lists such as Figure \@ref(fig:shopping-hand)  -- a dozen lines scrawled on the back of a used envelope stuck on the refrigerator door  --  and the five ranked items shown to a customer on an online shopping site (as shown in Table \@ref(tab:random-basket)). Even if they are similar in content,  the list of five suggestions shown to Tesco's online customers at the checkout  exemplifies a transformation in forms of ordering. In its re-writing of the list, the recommender system reassembles individualised desire, consumption practices, value-chains, infrastructural time-space dynamics, the lives of others and the propensities of things.   A major contemporary form of ordering, the recommender system can be found in many  forms of news, media, entertainment, marketing, and advertising. Tesco's recommender systems differs greatly from others associated with Amazon, Netflix or Ebay. Grocery shopping is a high-frequency, multi-purchase shopping practice and the potential for recommendations proliferates accordingly.

Despite being titled 'Have You Forgotten?' Tesco's recommender system  focuses on 'conversion events.' If a shopper does buy a recommended item, 'conversion' has occurred.  From a supermarket perspective, high-value conversion are desirable. There is little value in successfully recommending small or cheap items such as rubber bands or toothpicks. Conversion events, when they occur, confirm the power of predictive analytics to change the world in some way. From a sociological perspective, the significance of conversion events is far less obvious. In the name of conversion events, recommender systems stage an epistemic-operational transformation central to big data. In conversion events,  predictive analytics happens in the world.  The tendency of things to realize certain events -- what Popper terms the 'propensities of things' [@Popper_1990, 11] -- happens  through data practices.   Data in concert with calculation, models, infrastructures, and experiments, affords apprehension, visibility and management of associative propensities in the world.

# Lists and orders

In grocery shopping, conversion events derive from and affect lists. Whether  they write out shopping lists or not, online grocery shoppers interact with commodities through lists.  Lists have long interested historians, anthropologists, sociologists, literary and media theorists as everyday ordering devices. If, as the social anthropologist Jack Goody argued, lists are historically primary as forms of writing [@Goody_1986], and if list-making and its later variants (e.g. tables) precede discursive and narrative writing practices, then we might expect lists to function as powerful social ordering devices. More recent sociological work on lists (see [@deGoede_2016] for an overview) explore the social and political potency of lists as ordering devices. As literary scholars suggest (see [@Mainberger_2003]), even if lists have often been de-valued as literary forms, list-making commonly appears in literary form whenever writing seeks to address, name, group or evoke totality, profusion, excess or abundance. From both  anthropological and literary perspectives, shopping lists have excellent reasons to exist: they are deeply rooted in organisational life and infrastructures. Lists weave together people, infrastructures, things, and places.[^1.456]

In this paper, I focus on lists and list-making in terms of the predictive social orderings.   All lists are forms of order. Mathematicians distinguish lists from sets on the basis of order. Formally, both lists and sets are collections, but a list also contains a mapping of its elements to an ordering, usually the natural or counting numbers. Sometimes this order is explicit. Numbers are written down one side of the list. Often the order is implicit to the spatial arrangement of the list.  Mappings between list order and social order are manifestly highly contingent. A flight checklist differs greatly from a personal to-do list, even if they share many formal properties.  Listing practices in a given social setting, therefore, always index an engagement between the formal properties of the list and the necessarily more fluid dynamics of social order. From the perspective of social order, shopping lists, as they are co-constructed by a supermarket chain's recommender system, might be important intersectional zones for the conversion of average everydayness through data-analytic propensities.

How should we view the predictive re-writing of lists? The media theorist Mark Hansen has recently suggested that 'predictive analytics are discoveries of micrological propensities that are not directly correlated with human understanding and affectivity and that do not by themselves cohere into clearly identifiable events.' [@Hansen_2015a, 111-112] Hansen's account resonates with wider views of big data.  The expectation energising much predictive practice  pivots on 'micrological propensities' elicited from data. Unlike many other accounts of data, Hansen  links data to propensities of things in the world via probabilities:

>Whatever explanatory and causal value predictive analytics of large datasets have is, I suggest, ultimately rooted in this ontological transformation whereby probabilities are understood to be expressions of the actual propensity of things [@Hansen_2015a, 120]

In his Whitehead-inspired account, Hansen theorises prediction as an 'ontological transformation' that deploys probabilities in an evermore closely woven and encompassing expression of animated, eventful, propensities of things. Predictive analytics links probabilities as calculations to propensities or the mutable associative agencies of things. The practical and indeed empirical  question is whether such transformations or 'conversion' in relations between probabilities and 'the propensities of things'  can be detected and articulated in the prosaic setting of shopping lists and recommender systems.[^20.1]

[^20.1]: I will argue that 'conversion' is a preferable term for these changes.  It highlights  re-orientations in subjects, experience, things, numbers and structures that ranges more widely than Hansen's 'ontological transformation' formulation.

# Conversion events: DunnHumby's work for Tesco

@@

```{r simulate-inventory, echo=FALSE, cache=TRUE, fig.margin=TRUE, fig.cap='Prices and quantities of things in the Tesco universe of reference'}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)

csvs = list.files('data', pattern = '*.csv',full.names=TRUE)
df = bind_rows(sapply(csvs, read_tsv))

g = ggplot(df, aes(x=price,y=ContentsQuantity)) + geom_point(alpha=0.3, size=0.9 ) + theme(axis.text=element_text(size=10))  + xlab('Price (£)')
g + ylab('Number of products') + scale_x_log10() + geom_smooth()
```

```{r expensive-groceries, echo=FALSE, cache=TRUE}
s = sample_n(df[df$price>=30,], size=10)
s$Price = s$price
s$Label = s$name
kable(s[order(s$price), c('Label','Price')], col.names=c('Item', 'Price (£)'), caption='Random sample of higher cost items from Tesco')
```

Like many recommender systems, Patel's models address a large inventory of commodities.  Figure \@ref(fig:simulate-inventory) plots something of the density of commodity prices in the Tesco grocery inventory. By virtue of their sheer abundance, the plot cannot easily separate the many items that Tesco sells, especially in the price range £1-£10. A random sample of the more highly priced items in Tesco stores is shown in Table  \@ref(tab:expensive-groceries)

Against the background of the sheer number of commodities and their distribution of prices,  Patel's presentation presents two opportunities. First, by recounting, contextualising and commenting on the main steps in making the 'Have your forgotten?' list, we might follow some of the predictive sense-making done by data scientists and customer analytics teams working with transactional data in a typical commercial setting. Patel mentioned many of these steps only fleetingly in the presentation, for they are largely taken for granted as part of predictive analytic practice. Second, Patel focused on the renovation and updating of long-standing data-mining practice via a much more explicitly 'big data' and 'machine learning' oriented implementation.  What stands out from the presentation is the normality of the recommender system: it is part of a long-standing and ongoing transformation or conversion of grocery shopping. From a sociological standpoint, the interest lies less in specific technical innovations and more on how the ordering work done by recommender systems relates to problems of social order more generally.

Since Patel was not able to release her presentation slides to a wider audience, I rely on data-driven simulations and statistical programming as ways of reconstructing elements of the recommender system. In order to explore the operational environment she described, and its predictive operations, I model key features of her presentation using sample grocery data drawn from the Tesco API ([Application Programmer Interface](https://devportal.tescolabs.com/docs/services/56c73b1bf205fd0ed81dbe7a/operations/56d5b2bee2813e0a0053c47c), [@Tesco.com_2017])   and a range of data science, statistics and graphics libraries implemented in the widely-used statistical programming language `R` [@RDevelopmentCoreTeam_2010]. At a methodological level, it draws heavily on the vision of 'interface methods' proposed by Noortje Marres [@Marres_2015] in order to keep in view both the highly technical orderings of lists in recommender system and the changes in social orderings they might imply or impel.

# How the shopping list was reconstituted in the 1990s

@@

While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs or at least might want to continuously up-sell and cross-sell to customers who might hardly ever visit the supermarket itself.

How could we characterise the shift from _ClubCard_ data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [@Silverwood-Cope_2014](https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market), then Tesco itself has undergone some kind of conversion? Patel described her work at DunnHumby as shifting the recommender system from a 'rules-based list' to a 'relevancy model.' The relevancy model changed how the 'Have You Forgotten' list was constructed.   This shift is a typical of broader re-organisation of prediction in which existing data analysis practices are being re-distributed and intensified in particular ways. The shift in models results in a more probabilistic structuring of lists.  The ostensible banality of the 'Have you forgotten?' recommendations betrays micrological signs of a topologically complex predictive infrastructure that constructs alignments between people and things in significant ways.

# Reconfiguring recommendations

Academic researchers first began writing about recommender systems in the mid-1990s. From the outset they highlighted a potential shift from demographic-driven market research or data mining techniques to personalized recommendations. For instance, writing in 1997 in a special section of the _Communications of the ACM_ on recommender systems, Paul Resnick and Hal Varian (at that time Dean of Information Sciences at UC Berkeley, but currently Chief Economist at Google), made much of this personalisation [@Resnick_1997]. Resnick and Varian  emphasise the need to distinguish the emerging practices from data mining:

@@

```{r arules-ex, cache=TRUE, fig.cap='Frequency of association betwene items in the `Grocery` dataset', message=FALSE, echo=FALSE, fig.margin=TRUE}
library(knitr)
library(arules)
library(arulesViz)
library(datasets)
data(Groceries)
itemFrequencyPlot(Groceries,topN=10,col=rainbow(10), ntype="absolute",  cex.axis = 0.55)
```

```{r arules-ex2}
options(digits=2)
rules<- apriori(data=Groceries, parameter=list(supp=0.001,conf = 0.15,minlen=2), appearance = list(default="rhs",lhs="whole milk"), control = list(verbose=FALSE))
rules_sorted <- sort(rules, decreasing=TRUE,by="confidence")
kable(inspect(rules_sorted[1:5]), caption='The first five association rules for the `Groceries` dataset')
```

The `apriori` algorithm finds sets of items that commonly occur together in transactions.  These commonly occurring sets are expressed as 'association rules.' For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart *might* have discovered   the fabled association between diapers and beer.[^149.1] Figure \@ref(fig:arules-ex) shows how often common items appear. `Whole milk` appears most frequently.  The association rule-based model for the `Groceries` dataset illustrates some characteristic ordering practices of 1990s-style data-mining approaches to lists (the `Groceries` dataset  was acquired from a 'local German supermarket' [@Hahsler_2006a]). The association between items is expressed in the form of _rules_ not a model. These rules, a term derived from older decision support literature, define sets of associations between items. Table \@ref(tab:arules-ex2) shows that the first rule has a stronger association than the second. Milk and vegetables is more likely an association than milk and buns.  A rules-based recommender system should, therefore, might put them in that order in the 'Have You Forgotten' list.

[^149.1]: One of the founding narratives of big data customer analytics, the beer-diaper association was not, according to [@Power_2002] found through data mining at all.

Although association rules and the `apriori` algorithm present social order in terms of sets of frequency-weighted associations, an `apriori`-based recommender system contains no statistical model. It also no grasp on personal relevance since the associations it defines are only between things. Put in terms of the big data conversion, the rules-based system is not predictive, or personal. The `apriori` algorithm might be seen as an attempt to grasp the excess of things in a Tesco supermarket by identifying subsets whose closure derives from the associations between things that people construct (through buying habits, by virtue of a shopping list, etc.). In association rules, personhood or experience is irrelevant. The openness of these rules -- the `apriori` algorithm can product any number of rules of diminishing strength -- points towards the propensity of things to find themselves together in shopping baskets, but cannot cannot individuate that propensity beyond demographic quantiles and categories.

Even if `apriori`  elicits associations between things, it struggles with the propensity of commodities to multiply. A simple illustration of the combinatorial problem faced by recommender system can be developed using the `Grocery` dataset.  If we take all the items in the `Grocery` dataset and paste them into the 'shopping list' box on the Tesco grocery website (or as I side, run them as searches on the TescoLab Product API), each of the 169 items in the dataset yields dozens and sometimes thousands of products from Tesco online.

```{r tesco-ex, echo=FALSE, cache=TRUE, fig.cap ='Tesco surplus',  fig.margin=TRUE}
    library(ggplot2)
    tesco = read.csv('data/reference_data/tesco_groceries.csv')
    total = sum(tesco$actual, na.rm=TRUE) 
    tesco = na.omit(tesco)
    tesco$actual = as.integer(tesco$actual)
    ggplot(tesco[tesco$actual>50,], aes(x=labels, y=actual, fill=labels)) +geom_bar(stat='identity') + coord_flip() + ggtitle('Tesco grocery items with more than 50 products') + theme(legend.position="none", axis.text=element_text(size=7))
```

Items in the `Groceries` dataset proliferate into a Tesco's list of branded variations. The 169 items of the `Groceries` dataset  expand into roughly `r total` Tesco items (see Figure \@ref(fig:tesco-ex)).  There is, I would suggest, a logistic propensity in things that recommender systems encounter in the ways of global supply chains. The multiple listed  in the `Grocery` dataset becomes more open in this setting through new impersonal intersections that must be bounded and ordered. The proliferation of things on the shelves of supermarket or grocery warehouse confronts produces a combinatorial problem for data miners. `r total` products (actually Patel mentioned 200,000 products) can be combined in a vast number of ways. If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible lists. Some of this vast number of possible shopping lists are highly unlikely in terms of typical purchasing patterns, but potentially valuable as recommendations.

# Ordering relevance through personhood or thinghood?

Combinatorial surplus, however, does not exhaust what might be found in the data. The relevancy model that DunnHumby implemented for Tesco (as for so many other  businesses) seeks to balance the tensions between proliferating associations and relative closure. It responds to the problem of finding an ordering in which the propensities of things as commodities and the propensities of people mingle with each other, without overflowing  the site of their  ordering encounter, the list.[^181.1] Big data discourse in its promissory mode  attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. As Turow and co-authors conclude in their account of transformation of retail space: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public force' [@Turow_2015, 476].

Mark Hansen offers a different perspective on what might happen in these ordering encounters. Hansen attributes 'real potential' to data. Hansen attributes an associative propensity to data:

> [S]urrounding any delimited predictive system is a larger field of data -- what I elsewhere call a “surplus of sensibility” -- that, viewed speculatively, indexes the causal efficacy of the total situation within which this delimited system operates. (Effectively, the latter gains its reliability from closing off this larger surplus of sensibility, thereby transforming an always excessive propensity into a [provisionally] closed dataset.) Because it affords data that exceed whatever any given predictive system might include, the data of the world’s causal efficacy -- the data constituting its real potentiality -- always and in principle facilitates knowledge that cannot be restricted to any particular agenda [@Hansen_2015a, 126-7].

Many affirmative and critical analyses of big data and predictive analytics highlight the divisive social orderings implemented through personalization. For good reason, they attend to changes in personhood.  Hansen, by contrast, seeks to conceive of the potential of data to support or provoke different relations to the world. Even if the underlying contrast between an open world and a 'provisionally closed dataset' limits the account in ways that I will explore, the ontological transformation  Hansen describes  involves people, commodity-things, and infrastructures. The recommender system generates common codings for  the associative propensities of things, the logistical value-chains of supermarket commerce and  the forms of personhood associated with shopping. A model of personhood based on purchase data delimits the propensities of things.

[^181.1]: Rather than seeing predictive analytics as a closed form of an open world, I am suggesting that we might begin to locate it in the joint effectuations or intersections within models.

If we think of the  new model that Patel describes as the delimited (or delimiting?) system within   the larger surplus of sensibility, how would the idea of a predictive system as a provisional enclosure of the world's causal efficacy, its 'surplus of sensibility' (a terms that carries some connotations of the critical political economy of surplus value) help us understand what the 'Have you forgotten?' recommendations do? It enables us to understand -- as we will see next -- what the Tesco relevancy model borrows from the world. By contrast, despite its ontological ambitions, it limits our understanding of how predictive analytics  themselves exist amidst technical ensembles whose operation remains obscure yet central to contemporary worlds.

```{r models, fig.cap='Precision measurements for different machine learning models'}
include_graphics('figure/patel_compares_models.png')
```

Patel described the data that Tesco keeps on grocery purchases: 'We have 52 weeks of purchase data for each customer.' These transformations have been widely discussed in the context of big data, where questions of who has how much data are usually most prominent (see [@Kitchin_2014]). In terms of the ways that lists order the world, the amount of data is not perhaps not so important as the way in which list-making is re-ordered via a model of personhood. Patel introduced the new relevancy model with a graph familiar to machine learners and statistical modellers. (A sketch of her graph appears in Figure \@ref(fig:models).)  The graph plots the _precision_ -- the proportion of the recommended products that customers actually purchase -- for several different models. In Hansen's media theoretical terms, the graph indexes the 'causal efficacy' of the recommender system, its capacity to include and transform propensities or 'real potentialities' into operational knowledge.   The graph compares the old rule-based recommender systems with some of the relevancy-model alternatives -- logistic regression, random forests, and a few others -- in terms of their predictions and how those predictions turned out. Patel dismissed most of the models quite quickly and focused only on one, the logistic regression model, which did better than other alternatives. Why did she say little about the model, which after all, produces the probabilistic predictions on which everything in the recommendation list depends and hence the conversion experience? Her interest lay in the precision of the recommendations, and, crucially for our purposes, in how 'most of the gain comes from thinking of good features.'

The models Patel so briefly mentioned are all classifiers, predictive models that spatially order  variables in a dataset as a high dimensional vector space. Operating in the vector space, models classify people and events. Much of the operational formation of the classifier models received scant reference in Patel's presentation. It is likely that in DunnHumby's models the classifying model at the heart of the recommender system constructs a decision boundary in vector space between points that belong to different classes of things [@Mackenzie_2015e]. The classes of things are binary: `recommended` or `not recommended,` and for many of the models shown in the graph, membership of a class is expressed as a probability.  The logistic regression model generates probabilities for each product for each customer. Patel did not articulate the implication of the contrast between the left and right hand side of the graphic. Although the statistical framing appears explicitly (e.g. in measuring the precision of  recommender systems), the underlying shift to a statistical model is barely mentioned. In becoming predictive, the 'Have You Forgotten?' recommendations have included a version of the social in the form of a  model that seeks to identify relations between variables.  The relevance of recommendations pivots on the apprehension of social order by a model.

# Sparsity: output 20 recommendations from 200,000 products

Note that neither the data analytic infrastructure that processes  Tesco customer data  - Patel will mention `Hadoop` -- nor the predictive analytic device -- the logistic regression model -- loom very large in the conversion of the recommender system. 'Most of the gain comes,' Patel observes, 'from thinking of good features.' While this is almost a truism in big data practice (see, for instance, [@Domingos_2012]), what counts as a good feature in a shopping list?

Rather than a surplus of sensibility, predictive models often encounter a dearth of sensibility in the data. A  very brief comment from Patel -- 'we have lots of zeros' -- points to the difficult terrain of  'good feature' thinking: the vast, nearly empty matrix of customer-product associations. Given that most things in Tesco remain relatively unknown to each other and to customers. A matrix that records associations between individual people and different products is bound to be mostly empty. Say Tesco has 1 million online customers. Each online  shopper has bought some selection of the 200,000 products. The customer-product data matrix will be `r format(1e6 * 2e5)` in size.  The product-customer matrix, the basic vector-space in which all recommender systems operate, remains very sparse and unpopulated. Given that any one customer is likely to only have bought 100 or so different products, the probability matrix will be close to `r format(100-(100/2e5)*100, 2)`% empty. The data is, as Patel put it, 'massively unbalanced' (any dataset where the items of interest are much rarer than some normal values is said to be unbalanced) and such imbalance would heavily bias the recommender system towards common and only impersonally relevant suggestions, suggestions that might  not produce the desired conversion experience.

```{r sparsity-ex, echo=FALSE, cache=TRUE, fig.margin=FALSE, fig.cap ='Sparsity of the customer-product association matrix' }
library(ggplot2)
library(reshape2)
p = as.logical(rbinom(n=2e5, size=20,0.001 ))
m = (matrix(p, nrow=1000, byrow=TRUE))
melted <- melt(m)
melted$value <- as.logical(melted$value)
ggplot(melted, aes(x = Var2, y = Var1, fill = value)) + geom_tile() +
    scale_fill_manual(values = c("grey", "red")) + xlab('Customer id') + ylab('Product id') + theme(legend.position='none')
```

>>>>

# Probabilities in the data analytic conversion of social order

# A. Introduction  includes a,b,c from above

This paper is a philosophical reconstruction [@Dewey_2017] of an online grocery ordering system. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Given this highly mediated and limited encounter with a recommender system and one of its designers, the paper constructs a multi-layered account. It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping. It conducts several code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some important elements of the system in question.  It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map the functional statements and practices underpinning the knowledge and order generated by such systems over time. Not least, I draw on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations around probabilities and their enactments. Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Rabinow_2004; @Bogost_2012], through this combination of approaches,  the paper itself provisionally allows its own reconstruction to be shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the grocery ordering system, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture. Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _physical_ reality, we might come to understand the world they belong to as what the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] might shift.

The broad implication of the argument concerns how we think about social order or regularities: proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming organisations, institutions, economic processes and social life. I suggest too that a transformation or conversion event is occurring, but one which might open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh.  Even if the argument is centred on an eminently social practice such as shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- particle physics for instance -- do not.

# B. Patel section + opening description

@@

Against the background of the sheer number of commodities and their distribution of prices,  Patel's presentation presents two opportunities. First, by recounting, contextualising and commenting on the main steps in making the 'Have your forgotten?' list, we might follow some of the predictive sense-making done by data scientists and customer analytics teams working with transactional data in a typical commercial setting. Patel mentioned many of these steps only fleetingly in the presentation, for they are largely taken for granted as part of predictive analytic practice. Second, Patel focused on the renovation and updating of long-standing data-mining practice via a much more explicitly 'big data' and 'machine learning' oriented implementation. Her presentation concerns what we might term, for reasons that will gradually emerge, a 'conversion event' in which a long-standing recommender system was replaced by a new, explicitly 'big data'-style system concerned with 'personally relevant' recommendations.   What stands out from the presentation is the normality of the recommender system: it is part of a long-standing and ongoing transformation or conversion of grocery shopping. From a sociological standpoint, the interest lies less in specific technical innovations and more on how the ordering work done by recommender systems relates to problems of social order more generally.

Online grocery shopping at [Tesco](https://www.tesco.com/groceries/?icid=dchp_groceriesshopgroceries), the largest supermarket chain in the UK, includes recommendations for further grocery purchases under the title of 'Have you forgotten?'. When Tesco customers shop for groceries online, a list of five recommendations appear at the checkout stage. The recommendations are the product of a recommender system, an important category of operational device in big data (see for instance [@Striphas_2015] for analysis of Netflix recommendation; [@Morris_2015]  or [@Seaver_2015] for an account of music recommendations).  The question 'have you forgotten?' is followed by a list of some  grocery items that could have been or are usually on a shopping list. The title of the suggestions is a bit misleading: the recommender system, as we will see, is not concerned with forgetting, with the many slips and oversights associated to which shopping is prone,  but rather with items that customers had not selected perhaps because they had never thought of buying them in the first place.

The shelves of large contemporary supermarket are the endpoint of global logistic supply chains, in all their logistical, value-transforming and brand-mediated hypercomplexity (see [@Neilson_2012; @Tsing_2009]).  Groceries imply a planetary geography of agriculture, industry, transport, communication and financialisation animated by flows of labour and capital. Encounters between this hyper-complex commodity-geography and people, even in the familiar confines of a supermarket, are no simple matter, either for shoppers or for supermarket operators such as Tesco.

# C. Shopping  lists and  a theory of social ordering

The anthropologist  Daniel Miller has argued [@Miller_2012] that  shopping takes place as a negotiation of discrepancies between normative and actual social order (for instance, between ideals of health and commitments to thrift). What Miller suggests here is that ... [TBA]  It can also be seen  as a sacrificial ritual that re-instates social order in various ways.

From an ethnomethodological standpoint, shopping lists provide important clues to how local social order is constituted, maintained, and repaired. Social ordering is done in shopping, and the shopping list as an inscriptive device stabilises certain forms of ordering at the expense of other possible orderings.

The shopping list, whether written on the back of an envelope, or saved as a list in an online grocery shopping system, lies at the intersection of logistical flows, infrastructural orderings, and  lively negotiations around actual and normative social orders.  Shopping lists, therefore  are intersectional ordering devices that encapsulates a universe of possible references, and a teeming multitude of propensities with an actual local order (in both senses of that term: an ordering and an imperative). As people shop, either by  trawling along aisles packed with thousands of products, or scrolling down screens or searching for particular brands amidst search results, lists inscribe some kind of order that filters or reduces the excessive propensities, claims, dazzle and distraction of commodities to the practicalities of domestic economy.

If, as the social anthropologist Jack Goody argues, lists are historically primary as forms of writing in urbanizing cultures [@Goody_1986], and if list-making and its later variants (e.g. tables) precede discursive and narrative writing practices, then we might expect lists to function as powerful social ordering devices. More recent sociological work on lists (see [@deGoede_2016] for an overview) explore the social and political potency of lists as ordering devices. As literary scholars suggest (see [@Mainberger_2003]), even if lists have often been de-valued as literary forms, list-making commonly appears in literary form whenever writing seeks to address, name, group or evoke totality, profusion, excess or abundance. From both  anthropological and literary perspectives, shopping lists have excellent reasons to exist: they are deeply rooted in organisational life and infrastructures. Lists weave together people, infrastructures, things, and places.[^1.456]

Finally, and more formally, we might also consider some of the formal properties of lists as orderings. All lists imply order, but what kind of order? Mathematicians distinguish lists from sets on the basis of a distinction focused on order. Formally, both lists and sets are collections, but a list also contains a mapping of its elements to an ordering, usually the natural or counting numbers. Sometimes this order is explicit. Numbers are written down one side of the list. Often the order is implicit to the spatial arrangement of the list. This formal idea of the list as a mapping is helpful since it suggests relational possibilities. Mappings occasion relations between sets, collections, group or multiplicities.  Mappings between list order and social order are manifestly highly contingent. A flight checklist differs greatly from a personal to-do list, even if they share many formal properties.  Listing practices in a given social setting, therefore, always index an engagement between the formal properties of the list and the necessarily more fluid dynamics of social order. From the perspective of social order, shopping lists, as they are co-constructed by a supermarket chain's recommender system, might be important intersectional zones for the conversion of average everydayness through data-analytic propensities.

TBA - short paragraph summarising the social theory of list ordering

# D. Archaeology of recommendations  from 1984  to 2007

@@

While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs or at least might want to continuously up-sell and cross-sell to customers who might hardly ever visits the supermarket itself.

How could we characterise the shift from ClubCard data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [How SEO helps Tesco to dominate the online grocery market](https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market)[@Silverwood-Cope_2014], then Tesco itself has undergone some kind of conversion? Patel described her work at DunnHumby as shifting the recommender system from a 'rules-based list' to a 'relevancy model.' The relevancy model changed how the 'Have You Forgotten' list was constructed.   This shift is a typical of broader re-organisation of prediction in which existing data analysis practices are being re-distributed and intensified in particular ways. The shift in models results in a more probabilistic structuring of lists.  The ostensible banality of the 'Have you forgotten?' recommendations betrays micrological signs of a topologically complex predictive infrastructure that intervenes in list-based social ordering to reconfigure alignments between people and things.

This is not a drastic or abrupt event. Academic researchers first began writing about recommender systems in the mid-1990s. From the outset they highlighted a potential shift from demographic-driven market research or data mining techniques to personalized recommendations. For instance, writing in 1997 in a special section of the _Communications of the ACM_ on recommender systems, Paul Resnick and Hal Varian (at that time Dean of Information Sciences at UC Berkeley, but currently Chief Economist at Google), made much of this personalisation [@Resnick_1997]. Resnick and Varian  emphasise the need to distinguish the emerging practices from data mining:

@@

# E. Experimenting with probabilistic conversions

# F. Conclusion
